A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-07 introduction.tex
reformulations

The Parikh image, more broadly, is a characterisation of formal
languages in terms of their character counts. Given a
language~$\Language$ over an alphabet~$\{a_1, \ldots, a_k\}$, the
Parikh image is a set of $k$-dimensional vectors that contains some
vector~$\VectorLiteral{m_1, \ldots, m_k}$ if and only if the
language~$\Language$ contains a word in which each $a_i$ occurs $m_i$
times. It is a classical result that the Parikh image of every
context-free language (and, thus, also of every regular language) is a
semilinear set, i.e., Presburger-definable~\cite{parikh-theorem}.
In
fact, it is possible to compute an existential Presburger formula
describing the Parikh image of any \emph{context-free} language in
linear time~\cite{generate-parikh-image} in the size of the grammar
describing the language. For the special case of
regular languages, this result was also stated
in~\cite{muscholl-linear}.

@@

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Joint satisfiability of Parikh images modulo Presburger constraints,%
        width=0.9\linewidth]
        Given regular languages~$\Language_1, \ldots, \Language_n$
        over a common alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula~$\phi(x_1, \ldots, x_k)$, 
        decide if there is a word~$w \in \Language_1 \cap
        \cdots \cap \Language_n$ whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
}

The problem can be solved using the classical construction of Parikh
images~\cite{muscholl-linear,generate-parikh-image} by first computing
the product of finite-state automata accepting the
languages~$\Language_1, \ldots, \Language_n$, then extracting a
Presburger formula~$\psi$ describing the Parikh image of the product
automaton, and finally checking the satisfiability of $\phi \wedge \psi$.

@@

Secondly, the constructed Presburger formula~$\psi$ contains a linear
number of existential quantifiers in the size of the product
automaton, as well as complex Boolean structure, which is needed to
express the connectedness of paths considered in the construction.
Solving formulas of this kind generally tends to be taxing for
solvers~\cite{ostrich-plus}; in the present case, since the product
automaton itself is exponentially big, also formula~$\psi$ has
exponential size in the number of considered regular languages, and
can easily become too complex for today's Presburger arithmetic
solvers to handle.

@@

Addressing these concerns, we have developed a calculus for Parikh images of
intersections of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

@@

For our problem, the constraints are automata connectivity and linear
equations (Presburger formulas).
We use propagation to lazily enforce connectivity of automata,
removing transitions as they become unreachable during solving or are
eliminated by constraints in the Presburger formula. We
also branch on the presence or absence of transitions to
encourage propagation of the connectivity constraint. This allows
us to whittle down the automata
(including intermittent products) before moving on with potentially
 explosive product construction. The underlying observation
is that though the domain of our integer variables
is infinite, the number of simple paths
through the automata (and their product) is finite. In summary, the key ideas are:

@@

In summary, we contribute:
\begin{itemize}
    \item The \Calculus{} calculus to efficiently compute (a homomorphism on)
          the Parikh image of products of Parikh automata.
    \item Techniques to efficiently implement \Calculus{} in a modern
    automated theorem prover, including strategies for case splitting, clause
    learning, and constraint propagation for connectedness.
    \item The \Catra{} tool for solving such instances, containing an
    implementation of \Calculus{}, the over-approximation described
    in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model
    checker~\cite{nuxmv}.
    \item Experiments illustrating the performance of \Calculus{} on real-world
    examples from string solving, including \NrBenchmarks{} instances in a
    standardised format made available for future study.
\end{itemize}

@@

The problem of computing constraints on Parikh images over products of regular
languages under a given commutative homomorphism amounts to checking emptiness
of products of
Parikh automata. Parikh automata are regular automata extended with integer
counters with given increments and decrements for each transition, where we
allow checking a set of linear constraints on the final values of the counters
(but not their intermittent values) \cite{parikh-automata}. Parikh automata
without constraints on the final values on their registers are also sometimes
called cost-enriched automata, weighted automata, or counter automata, depending
on exact definitions and side constraints. The decision problem tackled in this
paper, determining the emptiness of an intersection of Parikh automata, was
shown to be PSPACE-complete~\cite{graph-queries}.

@@

A common approach to handling Parikh image computation is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Language(\Automaton_1) \cap \ldots \cap \Language(\Automaton_k))$
with the conjunction of the automata's Parikh maps, $\bigwedge_{i=1}^{k}
\ParikhMap(\Language(\Automaton_i))$~\cite{approximate-parikh}.
Due to the over-approximation,
this approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}.

Our calculus~\Calculus{} is similar in spirit to the work
of \citeauthor{symbolic-boolean-derivatives}, who tackle the exponential
blow-up resulting from Boolean combinations
of regular automata in SMT string solvers
through the use of symbolic derivatives~\cite{symbolic-boolean-derivatives}.
 The research by
 \citeauthor{symbolic-boolean-derivatives} does not consider Parikh images,
 however. In addition to presenting a decision procedure for lazily dispatching
constraints, we similarly also allow for symbolic labelling of automata to
handle large Unicode alphabets.

>>>>

The Parikh image, more broadly, is a characterisation of formal
languages in terms of their character counts. Given a
language~$\Language$ over an alphabet~$\{a_1, \ldots, a_k\}$, the
Parikh image is a set of $k$-dimensional vectors that contains some
vector~$\VectorLiteral{m_1, \ldots, m_k}$ if and only if the
language~$\Language$ contains a word in which each $a_i$ occurs $m_i$
times. It is a classical result that the Parikh image of every
context-free language (and, thus, also of every regular language) is a
semilinear set, i.e., Presburger-definable~\cite{parikh-theorem}.
In
fact, it is possible to compute an existential Presburger arithmetic formula
describing the Parikh image of any \emph{context-free} language in
linear time~\cite{generate-parikh-image} in the size of the grammar
describing the language. For the special case of
regular languages, this result was also stated
in~\cite{muscholl-linear}.

@@

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Joint satisfiability of Parikh images modulo Presburger arithmetic constraints,%
        width=0.9\linewidth]
        Given regular languages~$\Language_1, \ldots, \Language_n$
        over a common alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger arithmetic formula~$\phi(x_1, \ldots, x_k)$, 
        decide if there is a word~$w \in \Language_1 \cap
        \cdots \cap \Language_n$ whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
}

The problem can be solved using the classical construction of Parikh
images~\cite{muscholl-linear,generate-parikh-image} by first computing
the product of finite-state automata accepting the
languages~$\Language_1, \ldots, \Language_n$, then extracting a
Presburger arithmetic formula~$\psi$ describing the Parikh image of the product
automaton, and finally checking the satisfiability of $\phi \wedge \psi$.

@@

Secondly, the constructed Presburger arithmetic formula~$\psi$ contains a linear
number of existential quantifiers in the size of the product
automaton, as well as complex Boolean structure, which is needed to
express the connectedness of paths considered in the construction.
Solving formulas of this kind generally tends to be taxing for
solvers~\cite{ostrich-plus}; in the present case, since the product
automaton itself is exponentially big, also formula~$\psi$ has
exponential size in the number of considered regular languages, and
can easily become too complex for today's Presburger arithmetic
solvers to handle.

@@

Addressing these concerns, we have developed a calculus for Parikh images of
intersections of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string length. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

@@

For our problem, the constraints are automata connectivity and linear
equations (Presburger arithmetic formulas).
We use propagation to lazily enforce connectivity of automata,
removing transitions as they become unreachable during solving or are
eliminated by constraints in the Presburger formula. We
also branch on the presence or absence of transitions to
encourage propagation of the connectivity constraint. This allows
us to whittle down the automata
(including intermittent products) before moving on with potentially
 explosive product construction. The underlying observation
is that though the domain of our integer variables
is infinite, the number of simple paths
through the automata (and their product) is finite.
\emph{In summary, the key ideas are:}

@@

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently check the satisfiability
  of the Parikh image of an intersection of regular languages modulo
  Presburger arithmetic side conditions.
    \item Techniques to efficiently implement \Calculus{} in a modern
    automated theorem prover, including strategies for case splitting, clause
    learning, and constraint propagation for connectedness.
    \item The \Catra{} tool for solving such instances, containing an
    implementation of \Calculus{}, the over-approximation described
    in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model
    checker~\cite{nuxmv}.
    \item Experiments illustrating the performance of \Calculus{} on real-world
    examples from string solving, including \NrBenchmarks{} instances in a
    standardised format made available for future study.
\end{itemize}

@@

The problem of satisfying Parikh images over products of regular
languages, modulo Presburger arithmetic side conditions,
amounts to checking emptiness of products of
Parikh automata. Parikh automata are regular automata extended with integer
counters with given increments and decrements for each transition, where we
allow checking a set of linear constraints on the final values of the counters
(but not their intermittent values) \cite{parikh-automata}. Parikh automata
without constraints on the final values on their registers are also sometimes
called cost-enriched automata, weighted automata, or counter automata, depending
on exact definitions and side constraints. The decision problem tackled in this
paper, determining the emptiness of an intersection of Parikh automata, was
shown to be PSPACE-complete~\cite{graph-queries}.

@@

\citeauthor{muscholl-linear} and \citeauthor{generate-parikh-image}
define a closed-form description of the Parikh image of any regular
language as an existential Presburger arithmetic formula.
%
An approach to optimise the construction is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Language(\Automaton_1) \cap \ldots \cap \Language(\Automaton_k))$
with the conjunction of the Parikh images, $\bigwedge_{i=1}^{k}
\ParikhMap(\Language(\Automaton_i))$~\cite{approximate-parikh}.
Due to the over-approximation,
this approach is primarily useful for unsatisfiable instances,
and requires falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}. Our calculus~\Calculus{}, in a somewhat
similar but more fine-grained manner, utilises laziness to postpone or
avoid the most expensive steps in the computation of Parikh images of
the intersection of regular languages.

Our calculus~\Calculus{} is also similar in spirit to the work
of \citeauthor{symbolic-boolean-derivatives}, who tackle the exponential
blow-up resulting from Boolean combinations
of finite-state automata in SMT string solvers
through the use of symbolic derivatives~\cite{symbolic-boolean-derivatives}.
 The research by
 \citeauthor{symbolic-boolean-derivatives} does not consider Parikh images,
 however. In addition to presenting a decision procedure for lazily dispatching
constraints, we similarly also allow for symbolic labelling of automata to
handle large alphabets.
